A Case of Genetic Dilution in Improved Maize: Estimating Comparative Advantage of Adoption in Ethiopia

### 2022-01-27 paper.tex

\documentclass{article}

\title{Estimating Comparative Advantage in Improved Seed Adoption: Hybrid Maize in Ethiopia}

\usepackage{natbib}
\bibliographystyle{rusnat}
\usepackage{multicol}
\usepackage{multirow}
\usepackage{booktabs}
\usepackage{caption}
\usepackage{amsmath}
\usepackage{longtable}
\usepackage{graphicx}
\usepackage{float}

\usepackage{graphics}

\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\doublespacing

\begin{document}

\maketitle

%%%% I added the bullet points where I think they fit what we have as an introduction

\begin{itemize}
    \item Tim Weiss- hybrid subsidies
    \item is there such a thing as traditional seeds?

    \item Let's evaluate the effect of adoption in the context of Ethiopia
    \item find ambiguous results, Why?
    \item have access to DNA fingerprinting data
    \item what do they say about adoption?
    \item turns out that "adoption" isn't really a reliable concept/variable
    \item most farmers are "adopting" under pretty conservative purity definitions
    \item Let's see what happens with different varieties of seed, does adoption of that seed lead to positive yield effects?
    \item can't use these methods but newer data has access to newer varieties where we're relatively more confident
\end{itemize}

@@

%%% Why it is important
Food security is a major challenge in many East African countries, and Ethiopia is no exception, having historically struggled to provide an adequate and reliable food supply \citep{Ramakrishna2002-hv, Jaleta2018-oj}. Since the drought of 1984, Ethiopia has become increasingly reliant on maize, as of today, it constitutes the most widespread crop in the country.\footnote{FAO stats data shows the total area cultivated with Maize in 2019 is 25\% larger than the one cultivated with the next most widespread crop (sorghum), and 27\% and 139\% larger than the next two following crops, wheat and Barley, respectively.} As Figure \ref{fig:maize_yields} shows, there is an upward trend in the total area cultivated with maize since 1990, with average yields that have almost doubled during this period. These increases in yields have had enormous consequences in decreasing undernourishment, as its prevalence has dramatically decreased from from 47\% of the population in 2001 to 14\% in 2018 (The World Bank, 2021). Despite such generalized progress, food insecurity remains affecting more than half the population of Ethiopia (The World Bank, 2021), which already speaks about the heterogeneous progress across subpopulations\footnote{The country has also recently seen increases in food insecurity because of COVID-19, the conflict in Ethiopia’s Tigray region and the desert locust outbreak}.

The large disparities in yields and food security between the African continent and the rest of the developing world are often attributed to different adoption rates of green revolution technologies, such as improved seeds. Evidence has shown that the adoption of improved seeds leads to increases in agricultural yields (\citep{Carter2014-fm}), food security (\citep{Shiferaw2014-op}), and poverty reduction (\citep{Minten2008-tj}). Although critics of green revolution technologies have also claimed that these technological packages have not been as profitable for farmers in Africa as promised, given the high costs of improved seeds compared to the gains in productivity

Evidence has shown that the adoption of improved seed leads to increases in agricultural yields \citep{Carter2014-fm}, food security \citep{Shiferaw2014-op} and poverty reduction \citep{Minten2008-tj}. Despite such positive impacts, the low adoption levels of improved seeds varieties in Sub-Saharan Africa remains a puzzle.

%%%%% The puzzle 
Despite numerous technological breakthroughs in maize germplasm with increased yields and drought tolerance, adoption of improved seeds has remained a challenge, with only 13\% of maize farmers consistently using improved seeds varieties, and the lion share of farmers (62\%) not using improved seeds at all during the period where data is available. Moreover, there is a large percentage of farmers who adopt new varieties inconsistently limiting their opportunities to learn about their benefits, see Table \ref{tbl:Trajectories}.  Evidence has shown that the adoption of improved seed leads to increases in agricultural yields \citep{Carter2014-fm}, food security \citep{Shiferaw2014-op} and poverty reduction \citep{Minten2008-tj}. Despite such positive impacts, the low adoption levels of improved seeds varieties in Sub-Saharan Africa remains a puzzle.

The facts sub-Saharan farmers continue to use traditional farming techniques when more modern, higher-return agricultural technologies are available remains an empirical puzzle where several answers have been proposed. These include imperfections in credit markets \citep{Croppenstedt2003-pq}, property rights \citep{Place2000-el}, social learning \citep{Conley2010-ue,Foster1995-bz,Munshi2004-og}, lack of commitment devices \cite{Duflo2009-iv}, and high transportation costs that increase the cost of agricultural inputs \citep{Byerlee2013-qk}. This issue is becoming more important as even current varieties of improved seed are over 20 years old and their effectiveness in terms of yield improvement and disease resistance is waning \citep{Abate2015-rj}.

Multiple explanations have been proposed for observed adoption rates, including the type of technology promotion mechanisms used by governments, NGO’s, and businesses, the characteristics of the improved seeds and their adaptation to specific agro-climatic environments \citep{Bird2020-nt}, as well as growing conditions, such as the type of soil, input use, and other farmer characteristics \citep{Munshi2004-og}. The consensus in this literature is that the success of improved varieties depends on a range of factors that need to be better understood. Success of improved seeds should be accompanied by complementary conditions and inputs that would allow the full potential of these varieties to be realized. Specifically, extension services and on-farm field trials, seed variety characteristics and rainfall have been found to be crucial to the adoption of improved maize in Tanzania \citep{Kaliba2000-jh}. In the specific case of Ethiopia, labor, fertilizer use, farmers’ experience with extension packages, rainfall suitability and prices have been shown to increase the opportunity cost of not adopting improved wheat varieties \citep{Wale2006-bv}. Market access and the accessibility of extension services are also critical in the adoption of improved chickpea varieties in Ethiopia \citep{Verkaart2019-ol}. Other important determinants of heterogeneous effects include risk aversion \citep{Holden2016-vy}, farm size \citep{Ghimire2015-bd}, and credit constraints \citep{Simtowe2008-jn,Balana2020-hx}.

A different branch of the literature provides an alternative explanation for such low adoption rates, focusing on heterogeneous potential returns to adoption \citep{Suri2011-oi}. This literature suggests that even if returns to adoption are high, farmers may fail to adopt if they face low comparative advantage to adoption. Other methods used in the literature fail to account for this heterogeneity, admitting only the presence of time-constant and/or time-varying unobserved heterogeneities (depending on the data structure) for identification purposes \citep{Kassie2018-xn,Falco2011-rt}. These methods ignore the possibility of a farmer’s comparative advantage of using an improved technology over a traditional one as playing a central role in their adoption decision.

We rely on Suri’s method and its extension in \citep{Tjernstrom_Emilia_Dalia_Ghanem_Oscar_Barriga_Cabanillas_Travis_J_Lybbert_Jeffrey_D_Michler_and_Aleksandr_Michuda2020-bc} to account for such heterogeneous comparative advantages and evaluate how these are explained by climatic conditions. In Ethiopia, rainfall deviations from historical averages are found to be the main driver of heterogeneous effects for adopting agronomic packages \citep{Marenya2020-kb}. Similar experiences from Malawi show exposure to past droughts is found to increase the adoption of drought-tolerant maize among farmers that also receive subsidies \citep{Katengeza2019-af} and the dis-adoption of traditional local maize \citep{Holden2016-vy}.

@@

In this paper, we revisit the mechanism proposed by  \citep{Suri2011-oi} that justifies this lack of technology adoption on the existence of farmers heterogeneity. In particular, the fact that farmers with high net returns to the technology use new technologies while those with low returns do not. Adoption of a technology involves several dimensions such as market integration, access to supplementary inputs like fertilizer, as well agro-ecological considerations such as climate, altitude and soil quality. Ignoring the heterogeneity of these factors can hide the true impact of adoption, as different sub-populations may be impacted by the same technology in different, and even opposite ways. One way to tackle this problem, proposed by \citep{Suri2011-oi}, centers around quantifying the comparative advantage of a household to adopting improved seed varieties. Even when average returns are high, farmers may face heterogeneous returns based on their own, unobservable, comparative advantage in adopting new technologies. Using a correlated random coefficient model, Suri’s seminal paper finds that farmers’ lack of adoption is driven by insufficient net benefits to the technology stemming from poor infrastructure. Similar work in Ethiopia on improved chickpea varieties suggests that the economic returns to adoption plays a major role in the decision to adopt \citep{Michler2018-wk}.

%% Expand a bit on the methods
In our paper, we build on that literature by investigating hybrid maize seed adoption in Ethiopia, focusing on how heterogeneous returns drive the decision to use improved seed varieties. We implement two complementary estimation strategies. First, we implement at Group Random Coefficients estimation as originally devised by \cite{Suri2011-oi} and later expanded by \citep{Tjernstrom_Emilia_Dalia_Ghanem_Oscar_Barriga_Cabanillas_Travis_J_Lybbert_Jeffrey_D_Michler_and_Aleksandr_Michuda2020-bc}. The advantage of our estimation strategy is that we do not rely on identification from a valid instrument,
but modelling the heterogeneity directly through a projection of their returns to adoption on their observed
”trajectory” of adoption. This first estimation strategy uses data from the first three survey rounds that allow us to follow the same households across time (ESS1 to ESS3). The results show returns to adoption are consistently negative, a result that runs against the careful lab development of the improved seed varieties. To expand on this results, our second estimation strategy implements an endogenous switching regressions that exploits the last round of the survey (ESS4). The latest survey round is not a a continuation of the panel of the first three rounds, but has novel data on the genetic finger prints of the seeds used, that we exploit to understand why hybrid seed varieties seem not to increase returns when adopted. Varying how strict of a definition for hybrid seeds we use, starting from second-generation and recycled (SR1), and a more strict definition, which includes only new varieties (SR2), we find that  farmers that adopted a seeds with a less pure genetic profile would have had better returns by simply not adopting.

We put together the results from both estimation strategies to suggest that the negative return to adoption we find under the Group Random Coefficients estimation reflects the loose definition of hybrid seeds, where only a subset of farmers are using the purest genetic varieties. Considering that poorer farmers are the ones paying a premium for genetically diluted hybrids, this result has several policy implication about the need to foster a more inclusive policy when helping farmers to obtain the best performing seeds.

\begin{figure}[H]
    \centering
    \caption{Maize: Total area harvested and Average yields}    \includegraphics{results/figures/Maize_yields.pdf}

    \figcaption{Source: Authors' calculations using FAO stats data
    Note: Yield means the harvested production per ha for the area under cultivation}
    \label{fig:maize_yields}
\end{figure}

\section{Data}

The current study uses the Ethiopia Socioeconomic Survey (ESS), a part of the Agricultural Sample Survey (AgSS), a survey designed to obtain production estimates for the major crops. The survey is a join effort of the the Central Statistics Agency of Ethiopia, the World Bank and CGIAR (Consortium of International Agricultural Research Centers) \citep{kosmowski2020shining} that aims at collecting information from a representative sample of households in the  the most populous regions of the country. The ESS contains several questionnaires targeted at getting an understanding of the state of agriculture. This includes a survey that is conducted post-planting and then post-harvest. These surveys are more important to our analysis as it gives a close look to inputs used in agriculture, the types of seeds used (in various levels of detail) and yields collected post-harvest.

There were significant differences between the analysis carried out in ESS 1-3 and the most recent, ESS4. Most notably, the ESS4 is no longer a panel dataset made up of the households of ESS 1-3, but it includes important information on the DNA fingerprinting of seed type, with which we can calculate misclassification rates. This is important information as the first three waves of ESS only included self-reported hybrid maize status. Since our estimation strategy requires a balanced panel dataset of maize growers, our main analysis will include the first three waves of the ESS. We use the fourth wave later for validation, and to explore the impact of specific types of improved seeds.

Figure \ref{map:regions} shows the spatial distribution of the locations of the our households of study. As previously mentioned these are households that grow either improved or conventional maize and that are observed for all three rounds of the survey. Most households come from the northwest region of Amhara, Tigray and Oromiya which is consistent with where maize is mostly grown in Ethiopia \citep{Abate2015-rj}.

\begin{figure}
    \centering
    \caption{Spatial Distribution of Households Surveyed}
    \includegraphics{results/figures/map_hhids.pdf}
    \figcaption{Note: Green markers denote the locations of households}
    \label{map:regions}
\end{figure}

Table \ref{tbl:summary} shows a set of summary statistics for our sample. The mean parcel size for a household is about 0.15 Ha, with a maximum size of a little less than 1.5 Ha. Harvest labor is disproportionately skewed to family participation with around 20 days of harvest labor being conducted by family members and only about 2 days for hired labor. This is indicative of either malfunctioning labor markets, income constraints or high supervision costs. The distances to some key areas are also highlighted in Table \ref{tbl:summary}. On average households are around 13 kilometers away from an asphalt road and almost 60 kilometers away from the nearest market. This makes both the purchase of improved seed and the marketing of yields subject to transportation costs and a difficult enterprise. Fertilizer costs are on average 8,350 Birr (~178 USD) and with a large amount of the sample not using fertilizer at all. This is particularly important in the context of improved maize seed use as it requires fertilizer to be most effective. Only 5\% of the sample irrigates their crops making households susceptible weather shocks.

% \resizebox{0.7\textwidth}{!}{
{\small\tabcolsep=3pt  % hold it local
    \input{results/tables/summary}
}

The ESS has has both self-reported and crop cut yields available. This provides two sources of yield information, although self-reported yields seem to be overly large compared to either fresh or dried crop cut yields. Self-reported yields are both more susceptible to behavioral biases, but also to confusion caused by differences in assumed units. In general, though, the sample with available yield information cuts down the available sample immensely, which limits the power of the estimation. It isn't clear whether unavailable yields numbers is due to some sort of selection or whether crop cuts were only collected randomly from a few households.

@@

Defining what constitutes an improved seed is challenging because of the different characteristics of seeds. A crop seed is considered “improved” in Ethiopia if it was tested by breeders and evaluated to be superior to existing varieties (MoA, 2013, SPIA, 2020). In Ethiopia, the government controls most of the seed system. Local agricultural offices or extension agents assess the demands of different varieties, the MoA decides the seed quantities to be produced, and actors such as regional seed enterprises, seed companies, seed unions, are involved in the production of certified seeds; and seeds are primarily distributed by seed producer cooperatives (SPCs), which account for 37\% of total distribution in the country (SPIA, 2020).

An improved seed can be obtained directly from such distribution centers (which would distribute a first generation seed), or it could be cultivated from second or posterior generations from the originally bred seed. Farmers are asked to report whether the variety they used was a first-generation improved seed, second generation,  recycled, or a traditional variety. When buying seeds, farmers might be able to identify the variety they are looking for, but it is unlikely that they would be aware of the source.

Fifty four varieties have been released in Ethiopia since 1990, out of which seven were released between 1990 to 1999, 29 between 2000 to 2009, and 18 between 2010 to 2019. Thirty four of such varieties contain CIMMYT germplasm, out of which, 10 Drought-tolerant maize varieties (DTMZ) and 8 Quality protein maize (QPM) were released (most of these between 2010 to 2019). Out of the CIMMYT-related germplasm, ten are Open-pollinated varieties (OPVs) and 25 are hybrid. 
The fourth wave of the ESS introduced for the first time DNA fingerprinting for barley, maize, and sorghum. The collection of crop-cuts for crop yield estimation facilitated the process for DNA fingerprinting. For the crop-cuts, a 4 square meter quadrant was randomly selected in each plot, so that total production harvested from such patch was weighted (both when freshly harvested and when dried). The DNA material was extracted from samples from such dried crop-cuts, that were then matched with the genetic material of the maize DNA reference library for Ethiopia, obtained from a CIMMYT and EIAR project. The collection of DNA samples focused only on eight regions. In each EA of such regions, a random sample of plots was selected, of a maximum of 10 fields for each crop. As a result, 505 plots were selected for DNA fingerprinting, which corresponded to 447 households.

>>>>

\documentclass[11pt]{article}

\usepackage{natbib}
\bibliographystyle{abbrvnat}
\setcitestyle{authoryear,open={(},close={)}} 
\usepackage{multicol}
\usepackage{multirow}
\usepackage{booktabs}
\usepackage{caption}
\usepackage{amsmath}
\usepackage{longtable}
\usepackage{graphicx}
\usepackage{float}
\usepackage{bm}
\usepackage{graphics}
\usepackage{threeparttable}
\usepackage{amssymb}
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\usepackage{afterpage}
\doublespacing
\usepackage{subfig}
\usepackage{epigraph}

\usepackage{enumerate}
\usepackage{tabularx}
\usepackage{fancyhdr}
\usepackage{lscape}
\usepackage{mdframed}
\usepackage{mathtools}
\usepackage{array}
\usepackage{import}
\usepackage{fullpage}

\begin{document}
    \begin{titlepage}
    \title{A Case of Genetic Dilution in Improved Maize: Estimating Comparative Advantage of Adoption in Ethiopia}
% This second title is my preferred title at this point (28 Oct 2021). Open to other titles, but it shouldn't be longer than this one!

%     \author[1]{Oscar Barriga-Cabanillas}  
%     \author[2]{Cristina Chiarella}    
%     \author[2]{Juan Sebastian Correa}  	
%     \author[3]{Aleksandr Michuda}

% 	\affil[1]{\small \emph{The World Bank}}
% 	\affil[2]{\small \emph{FAO}}
% 	\affil[3]{\small \emph{Assistant Research Professor at the Center for Data Science for Enterprise and Society, Cornell University}}
\author{%
 Aleksandr Michuda \footnote{Center for Data Science for Enterprise and Society, Cornell University}%
 \and Cristina Chiarella\footnote{UC Louvain}%
 \and Oscar Barriga-Cabanillas\footnote{The World Bank}%
  \and Juan Sebastian Correa \footnote{FAO}%
  }
%   Assistant Research Professor at the 
	\date{\today}

    \clearpage\maketitle
	\thispagestyle{empty}
	\vspace*{-2em}
    \begin{center}\begin{abstract}
			\noindent  
			Despite large investments in breeding and making high-yielding maize varieties available, evidence shows that few farmers consistently use them over time. We use an innovative methodology that studies the adoption trajectories of farmers over a three-round panel data set to explain low adoption rates with heterogeneous returns to adoption. By relying on self-reported use of improved seeds, we find that farmers do not adopt because they have a low comparative advantage and seemingly small to negative returns to adoption. To reconcile this result, we exploit unique information on maize DNA-fingerprinting collected over the same areas in 2018. We find that the negative returns to adoption mask heterogeneity driven by the variety of improved seeds used. Positive returns to adoption are found for those farmers that use higher-purity germplasm, drought-tolerant maize, and newly released varieties. Our findings point to the wide dispersion of older and genetically diluted varieties, for which poorer farmers are paying a premium. The implications of our findings speak to the need for policies to better target context and geography, expand accessibility of improved seeds, and make higher yields more inclusive.

			%Despite the large investments on creating better performing maize varieties, very few farmers consistently use them. We rely on four rounds of the ESS survey to answer this question. We find that relying on self reported use of hybrid seeds suggest there are negative returns to adoption. To reconcile this result, we use the information on genetic fingerprinting of seed available on the last ESS survey round to show that positive return for adoption appear only to exits for those farmers using seed with pure germoplast. We discuss the implications of our findings, showing that poorer farmers are the ones that consistently rely on second and third generation seeds, whose performance is debatable. 
        \end{abstract}\end{center}

        {\small \noindent\emph{JEL Classification}:
        	\emph{Keywords}: Improved maize, DNA-fingerprinting, Group Random Coefficient, Endogenous Switching Regression, Heterogeneous returns}

    \end{titlepage}
\maketitle
\newpage

%\begin{itemize}
%    \item Tim Weiss- hybrid subsidies
%    \item is there such a thing as traditional seeds?
%    \item Let's evaluate the effect of adoption in the context of Ethiopia
%    \item find ambiguous results, Why?
%    \item have access to DNA fingerprinting data
%    \item what do they say about adoption?
%    \item turns out that "adoption" isn't really a reliable concept/variable
%    \item most farmers are "adopting" under pretty conservative purity definitions
%    \item Let's see what happens with different varieties of seed, does adoption of that seed lead to positive yield effects?
%    \item can't use these methods but newer data has access to newer varieties where we're relatively more confident
%\end{itemize}

@@

%%% Why it is important
Food security is a major challenge in many East African countries, and Ethiopia is no exception, having historically struggled to provide an adequate and reliable food supply \citep{Ramakrishna2002-hv, Jaleta2018-oj}. Since the drought of 1984, Ethiopia has become increasingly reliant on maize, and as of today, maize constitutes the most widespread crop in the country.\footnote{FAO statistics show the total area cultivated with Maize in 2019 is 25\% larger than the next most widespread crop (sorghum), and 27\% and 139\% larger than the next two following crops, wheat and barley, respectively.} As Figure \ref{fig:maize_yields} shows, there is an upward trend in the total area cultivated with maize since 1990, with average yields almost doubling during this period. This increase in maize yields has also coincided with a dramatic decrease in undernourishment which has fallen from 47\% of the population in 2001 to 14\% in 2018 (The World Bank, 2021). Despite these positive gains, food insecurity still affects more than half the population of Ethiopia (ibid), reflecting that the progress achieved 
is not necessarily homogeneous across all subpopulations.\footnote{The country also has recently seen increases in food insecurity because of COVID-19, the conflict in Ethiopia’s Tigray region and the desert locust outbreak.}

\begin{figure}[htp]
    \centering
    \caption{Maize: Total area harvested and average yields} \label{fig:maize_yields}
    \includegraphics[width=.6\textwidth]{results/figures/Maize_yields.pdf}
    \vspace*{-1em}
    \begin{table}[H]
        \centering
        \begin{tabular}{p{0.6\textwidth}} 
            \begin{tablenotes}[flushleft]
                  \small
                  \item Source: Authors' calculations using FAOSTAT data 
                  \item Note: Yields are measured as the harvested production per ha for the area under cultivation
            \end{tablenotes}
        \end{tabular}
    \end{table}       
\end{figure}

%%%%% The puzzle 
The large disparities in yields and food security between the African continent and the rest of the developing world are often attributed to different adoption rates of green revolution technologies, such as improved seeds. Evidence has shown that the adoption of improved seed leads to increases in agricultural yields \citep{Carter2014-fm}, food security \citep{Shiferaw2014-op}, and poverty reduction \citep{Minten2008-tj}. Nevertheless, critics of green revolution technologies have also claimed that these technological packages have not been as profitable for many farmers in Africa as promised (with the exception of Ethiopia), given the high costs, and their heavy dependence on input subsidies \citep{Wise20}. \par
Even if the net returns to adoption of improved seeds are more conservative than often thought, the evidence regarding increased yields over traditional varieties is consistent across multiple contexts. Despite this evidence and numerous technological breakthroughs in maize germplasm with increased yields and drought tolerance, in the period between 2011 and 2015 only 13\% of maize farmers in Ethiopia have adopted improved seeds consistently, a large percentage has adopted new varieties inconsistently, and the lion share of farmers (62\%) have not used improved seeds at all during the same period.

Several answers have been proposed to explain such low adoption of improved varieties. These include mainly the lack of liquidity and other economic constraints \citep{Carter2014-fm}, growing conditions such as the type of soil, input use, and farmer characteristics \citep{Munshi2004-og}, imperfections in credit markets \citep{Croppenstedt2003-pq}, ambiguously defined property rights \citep{Place2000-el}, lack of commitment devices \cite{Duflo2009-iv}, high transportation costs that increase the cost of agricultural inputs \citep{Byerlee2013-qk}, and improved seed supply constraints \citep{Bird2020-nt}. In this context, there is still a large percentage of the population not using improved seed varieties, or using varieties that are over 20 years old. With such old seed, there is evidence that yield improvement and disease resistance in these seeds are waning\citep{Abate2015-rj}. The variety of these constraints and their dispersion across subpopulations and locations speaks to considerable heterogeneous returns to adoption.

@@

In this paper, we revisit the mechanism proposed by \citep{Suri2011-oi}, where a lack of adoption of improved seed varieties is the result of farmers' heterogeneous returns to the new technology. Ignoring such heterogeneity can hide the true impact of adoption, as different sub-populations' returns may be impacted by the same technology in different, and even opposite ways, despite it providing high average returns. The heterogeneous returns to adoption are modeled as a farmer's unobservable comparative advantage using a Correlated Random Coefficient (CRC) model. In the original paper, \citep{Suri2011-oi} finds that farmers’ lack of adoption is driven by insufficient net benefits to the technology stemming from poor infrastructure that increase the cost of access. Similar work in Ethiopia on improved chickpea varieties suggests that the economic returns to adoption plays a major role in the decision to adopt \citep{Michler2018-wk}.

%% Expand a bit on the methods
We build on this literature by investigating hybrid maize seed adoption in Ethiopia, focusing on how heterogeneous returns drive the decision to use improved seed varieties. We implement two complementary estimation strategies. First, we implement a Group Random Coefficients estimation, which builds on \cite{Suri2011-oi} in \cite{Tjernstrom_Emilia_Dalia_Ghanem_Oscar_Barriga_Cabanillas_Travis_J_Lybbert_Jeffrey_D_Michler_and_Aleksandr_Michuda2020-bc} using a General Method of Moments (GMM) Estimator that allows us to test the assumptions assumed by \cite{Suri2011-oi}, as well as provide more economically interpretable results. The advantage of our estimation strategy is that we do not rely on identification from a valid instrument, but a strategy similar to \cite{Chamberlain1984-uk}. For this strategy, we rely on the first three survey rounds of the Ethiopian Socioeconomic Survey (ESS), panel dataset that allows us to follow the same households across time (from 2011 to 2015). Using self-reported information on the seed variety (improved or traditional), we compare the adoption trajectories over three rounds. The results show that the returns to adoption are in most cases, not statistically significant, and only significant with wave 3 adoption. This posits a puzzle that motivates our next estimation strategy.

We expand our results using the fourth round of the ESS (conducted in 2018), which provides a unique opportunity to understand why improved seed varieties seem not to increase returns when adopted. The ESS4 collected DNA fingerprints of harvested cropcuts in selected plots and compared its genetic material with a reference library of available varieties in Ethiopia. This provides critical information not only about the seed types but also about their purity, sources, and their year of release. We exploit this uniquely rich information to compare different definitions of improved seeds. Specifically, we compare the returns to the self-reported use of new, second generation or recycled varieties, with those of improved seeds as defined by their genetic material, comparing different purity thresholds (or degrees of no-contamination with other varieties), year of release, types of varieties, and their sources. Because the latest survey round is not a continuation of the first three rounds, our second estimation strategy implements an Endogenous Switching Regression (ESR) approach, a standard estimation strategy to study the question of adoption with cross-sectional data (\citealt{Marenya2020-kb}, \citealt{Shiferaw2014-op}), given its capacity to account for unobserved heterogeneity in adoption.

The results of the ESR find negative effects of adopting improved seeds, as per farmer self-reports. But the comparison of the different improved seed definitions using the DNA fingerprinting sheds light on possible explanations for this apparent puzzle. We find that when the definition of improved seed is based on greater purity of the germplasm (95\% of purity or more), the returns to adoption for those that used such improved varieties becomes positive and significant. We obtain similar results when evaluating the specific effects of adopting drought-tolerant maize varieties, CGIAR or Exotic varieties, or more recently released varieties.

Taken together, the negative and not statistically significant returns to adoption we find from both estimation strategies reflect the loose definition of improved seeds when such improved varieties are self-reported by farmers. Only a subset of farmers use the purest genetic varieties, drought-tolerant varieties, or newly released varieties, which tend to provide higher returns. This implies that poorer farmers are paying a premium for genetically diluted hybrids, which has far-reaching policy implications about the need to foster a more inclusive policy when making the best performing seeds available to all farmers.

Our results contribute to the literature on the drivers of low adoption rates. The consensus in this literature is that the success of improved varieties depends on a range of factors, and that the widespread adoption of  improved seeds is the result of complementary conditions and inputs that allow farmers to extract the full potential of these varieties. Specifically, extension services and on-farm field trials, seed variety characteristics and rainfall have been found to be crucial to the adoption of improved maize in Tanzania \citep{Kaliba2000-jh}. In the specific case of Ethiopia, labor, fertilizer use, farmers’ experience with extension packages, rainfall suitability and prices have been shown to increase the opportunity cost of not adopting improved wheat varieties \citep{Wale2006-bv}. Market access and the accessibility of extension services are also critical in the adoption of improved chickpea varieties in Ethiopia \citep{Verkaart2019-ol}. Other important determinants of heterogeneous effects include risk aversion \citep{Holden2016-vy}, farm size \citep{Ghimire2015-bd}, and credit constraints \citep{Simtowe2008-jn,Balana2020-hx}.

% A different branch of the literature provides an alternative explanation for the low adoption rates, focusing on heterogeneous potential returns to adoption \citep{Suri2011-oi}. This literature suggests that even if returns to adoption are high, farmers may fail to adopt if they face low comparative advantage to adoption. Other methods used in the literature fail to account for this heterogeneity, admitting only the presence of time-constant and/or time-varying unobserved heterogeneities (depending on the data structure) for identification purposes \citep{Kassie2018-xn,Falco2011-rt}.

We contribute to this literature by exploring heterogeneous comparative advantages to adoption in Ethiopia between 2011 to 2015, controlling for important drivers of adoption such as rainfall (found to be the main driver of heterogeneous effects for adopting agronomic packages \citealt{Marenya2020-kb}, \citealt{Katengeza2019-af}), labor and input use. Our paper also provides a novel way of visualizing GRC results, which become difficult to interpret with longer time variables. Our findings also provide an alternative hypothesis to the low-returns to adoption, namely that the wide dispersion of older generations and recycled improved seeds, which may be both resulting in lower yields and altering farmers' input use in response to a mis-classified seed. This also implies a methodological discovery when investigating hybrid seed adoption: the absence of a true control group to use for analysis. With more time and more genetic dilution, it becomes more and more likely that even those that do report hybrid use are still likely using some level of hybrid.

The article continues as follows. In Section \ref{sec:data}, we describe the data used, the sample frame, general farmer and farm characteristics, and improved seed adoption levels. Section \ref{sec:strategy} presents the methodological details of our two estimation strategies. Section \ref{sec:results} shows the estimations results, and Section \ref{sec:conclusion} presents the conclusion and implications of our analysis.

\section{Data}\label{sec:data}

This study uses the Ethiopia Socioeconomic Survey (ESS), part of the Agricultural Sample Survey (AgSS), a nationally representative survey designed to collect agricultural information for the major crops in Ethiopia. The survey is a joint effort of the the Central Statistics Agency of Ethiopia, the World Bank and CGIAR (Consortium of International Agricultural Research Centers) \citep{kosmowski2020shining}. The ESS contains four survey modules designed to obtain specific information about the different stages in the agricultural season. A post-planting questionnaire obtains information about the parcel, crop roster, seed roster and input use. A post-harvest questionnaire collects information about crop harvests by parcel (which includes cropcut measurements for selected plots), and crop disposition. These two instruments are complemented with a household and a community questionnaire, which obtain information on household characteristics, labor, land, expenditure, and community characteristics such as market prices and economic activities, respectively.

\subsection{The Three Round Panel}

There were significant differences between the analysis carried out in ESS 1-3 and the most recent, ESS4. Most notably, the ESS4 is no longer a panel dataset made up of the households of ESS 1-3, but it includes important information on the DNA fingerprinting of seed type, which can be used to calculate misclassification rates. This is important information as the first three waves of ESS only included self-reported hybrid maize status. Since our estimation strategy requires a balanced panel dataset of maize growers, our main analysis includes the first three waves of the ESS. We use the fourth wave for validation, and to explore the impact of specific types of improved seeds.

Figure \ref{map:regions} shows the spatial distribution of the locations of the households in the study. These are households that grow either improved or conventional maize and that are observed for the three rounds of the survey. Most households come from the northwest region of Amhara, Tigray and Oromiya which is consistent with where maize is mostly grown in Ethiopia \citep{Abate2015-rj}.

\begin{figure}[H]
    \centering
    \caption{Spatial Distribution of Households Surveyed}\label{map:regions}
    \includegraphics[width=.7\textwidth]{results/figures/map_hhids.pdf}
    \vspace*{-1em}
    \begin{table}[H]
        \centering
        \begin{tabular}{p{0.6\textwidth}} 
            \begin{tablenotes}
                  \small
                  \item Note: Red markers denote the locations of households
            \end{tablenotes}
        \end{tabular}
    \end{table}  
\end{figure}

Table \ref{tbl:summary} shows a set of summary statistics for the sample. The mean parcel size for a household is about 0.15 Ha, with a maximum size of a little less than 1.5 Ha. Harvest labor is disproportionately skewed to family participation with around 20 days of harvest labor being conducted by family members and only about 2 days for hired labor. This is indicative of either malfunctioning labor markets, income constraints or high supervision costs. The distances to some key areas are also highlighted in Table \ref{tbl:summary}. On average, households are around 13 kilometers away from an asphalt road and almost 60 kilometers away from the nearest market. This makes both the purchase of improved seed and the marketing of yields subject to transportation costs and a difficult enterprise. Fertilizer costs are on average 8,350 Birr ($\thicksim$178 USD), with a large amount of the sample not using fertilizer at all. This is particularly important in the context of improved maize seed use as it requires fertilizer to be most effective. Only 5\% of the sample irrigates their crops making households susceptible to weather shocks.

\afterpage{{\footnotesize\tabcolsep=1pt  % hold it local
    \input{results/tables/summary}
    \vspace*{-4em}
    \begin{table}[H]
        \centering
        \begin{tabular}{p{0.7\textwidth}} 
            \begin{tablenotes}
                  \small
                  \item Note: Author's calculations using ESS. Maize farmers only.
            \end{tablenotes}
        \end{tabular}
    \end{table}  
    \newpage
}}

The ESS has both self-reported and crop cut yields available. This provides two sources of yield information, although self-reported yields seem to be overly large compared to either fresh or dried crop cut yields. Self-reported yields are both more susceptible to behavioral biases, but also to confusion caused by differences in assumed units. In general, though, the sample with available cropcut yields cuts down the available sample significantly, given that cropcuts were only collected for a subsample of randomly selected households, which limits their power.

@@

A crop seed is considered “improved” in Ethiopia if it was tested by breeders and evaluated to be superior to existing varieties  (\citep{MoA13}; \citep{kosmowski2020shining}). In Ethiopia, the government controls most of the seed system, with local agricultural offices or extension agents assessing the demands of different varieties. The MoA decides the seed quantities to be produced, and actors such as regional seed enterprises, seed companies, seed unions, are involved in the production of certified seeds, that are primarily distributed by seed producer cooperatives (SPCs), which account for 37\% of the total distribution in the country (\citep{kosmowski2020shining}). An improved seed can be obtained directly from such distribution centers (which would distribute a first generation seed), or it could be cultivated from second or later generations from the originally bred seed. In the survey, farmers are asked to report whether the variety they used was a first-generation improved seed, second generation, recycled, or a traditional variety. When buying seeds, farmers might be able to identify the variety they are looking for, but it is unlikely that they would be aware of the source.

Fifty four varieties have been released in Ethiopia since 1990, out of which seven were released between 1990 to 1999, 29 between 2000 to 2009, and 18 between 2010 to 2019. Thirty four such varieties contain International Maize and Wheat Improvement Center (CIMMYT) germplasm, out of which, 10 Drought-tolerant maize varieties (DTMZ) and 8 Quality protein maize (QPM) were released (most of these between 2010 to 2019). Out of the CIMMYT-related germplasm, ten are open-pollinated varieties (OPVs) and 25 are hybrid (\citep{kosmowski2020shining}).

The fourth wave of the ESS introduced DNA fingerprinting of barley, maize, and sorghum for the first time. The collection of cropcuts for crop yield estimation facilitated the process for DNA fingerprinting. For the cropcuts, a 4 square meter quadrant was randomly selected in each plot, and total production harvested from the quadrant was weighed (both when freshly harvested and when dried). The DNA material was extracted from samples from the dried cropcuts, and were then matched with the genetic material of the maize DNA reference library for Ethiopia, obtained from a CIMMYT and Ethiopian Institute of Agricultural Research (EIAR) project. The collection of DNA samples focused on eight regions. In each enumeration area (EA) of these regions, a random sample of plots was selected, with a maximum of 10 fields for each crop. As a result, 505 plots were selected for DNA fingerprinting, corresponding to 447 households.
